WZW-модели и теория Черна-Саймонса

### 2010-05-12 GeneralOverview/WZW-CS.tex

\documentclass[a4paper,12pt]{article}
\usepackage[unicode,verbose]{hyperref}
\usepackage{amsmath,amssymb,amsthm} \usepackage{pb-diagram}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage[russian]{babel}
\usepackage{cmap}
\usepackage{graphicx}
\pagestyle{plain}
\theoremstyle{definition} \newtheorem{Def}{Definition}
\begin{document}
\tableofcontents
\section{String theory}
\label{sec:string-theory}

Free relativistic particle action is equal to length of trajectory. 
\begin{equation}
  \label{eq:1}
  \begin{split}
  x^{\mu}(0)=x^{\mu}_0,\; x^{\mu}(1)=x^{\mu}_1\\
  S=\int_{0}^{1} \sqrt{\left(\frac{dx^{\mu}}{d\tau}^2\right)}d\tau
\end{split}
\end{equation}
There is Poincare invariance $x^{\mu}\to A^{\mu}_{\nu}x^{\nu}+ B^{\mu}$ and reparametrisation invariance $\tau\to f(\tau),\; x^{\mu}(\tau)\to x^{\mu}(f(\tau))$.

String action is the area of world-sheet:
\begin{equation}
  \label{eq:2}
  \begin{split}
    X^{\mu}(x^1,x^2),\; \mu=1\dots d\\
    S_N [X^{\mu}(x^1,x^2)]=\int \sqrt{det \frac{\partial X^{\mu}}{\partial x^a}\frac{\partial X^{\mu}}{\partial x^b}} dx^1 dx^2
  \end{split}
\end{equation}
This action is called Nambu action, and it is invariant under Lorenz transformations and reparametrisations.

Transition amplitude for string is given by following formula:
\begin{equation}
  \label{eq:3}
  A=\int {\cal D}X^{\mu}(x^1,x^2)e^{-S_N[X^{\mu}(x^1,x^2)]}
\end{equation}
The problem here is the definition of measure of integration. 
It is not strictly defined in most cases, so there is nothing unusual here.

We want to change ``integration variable'' in order to fix the arbitrariness connected with reparametrisations of world-sheet.

Obviously our theory should be invariant under reparametrisation:
\begin{equation}
  \label{eq:4}
  \begin{split}
    x^a \to f^a (x^1,x^2)\\
    X^{\mu}(x^1,x^2)\to \tilde{X}^{\mu}(x^1,x^2)=X^{\mu}(f^1(x^1,x^2),f^2(x^1,x^2))
  \end{split}
\end{equation}

But Nambu action is not invariant. So we will use equivalent Polyakov action, which has following form:
\begin{equation}
  \label{eq:5}
  \begin{split}
    S_P [g^{ab}(x^1,x^2), X^{\mu}(x^1,x^2)]=\int g^{ab}(x)\partial_a X^{\mu} \partial_b X^{\mu}\sqrt{g} \;d^2 x\\
    g=\det g_{ab}
  \end{split}
\end{equation}

Polyakov and Nambu actions are equivalent in the sense that equations of motion are the same after the exclusion of auxiliary field $g_{ab}$.
\begin{equation}
  \label{eq:6}
  \delta_{X}S_N=0 \Longleftrightarrow \left\{
      \begin{aligned}
        \delta_X S_P=0\\
        \delta_g S_P=0
      \end{aligned}
      \right.
\end{equation}
Polyakov action is invariant under Lorenz transformations and reparametrisations:
\begin{equation}
  \label{eq:7}
  \begin{split}
    x^a \to f^a (x^1,x^2)\\
    X^{\mu}(x^1,x^2)\to \tilde{X}^{\mu}(x^1,x^2)=X^{\mu}(f^1(x^1,x^2),f^2(x^1,x^2))\\
    g_{ab}(x)\to \tilde{g}_{ab}(x)=\frac{\partial f^{a_1}}{\partial x^{a}}\frac{\partial f^{b_1}}{\partial x^b} g_{a_1 b_1}(x)
  \end{split}  
\end{equation}

There is also one additional invariance, which is called conformal or Weyl invariance.
\begin{equation}
  \label{eq:8}
  \begin{split}
    g_{ab}(x)\to e^{\sigma(x)}g_{ab}(x)\\
    g\to e^{\sigma} g\\
    g^{ab}\to e^{-\sigma} g^{ab}\\
    S_P[e^{\sigma(x)} g_{ab}(x),X^{\mu}] = S_P[g_{ab}(x),X^{\mu}] 
  \end{split}
\end{equation}

So we have rather big freedom. For example, any metric on sphere can be obtained from one (constant) metric with reparametrisation and Weyl transformation.
\begin{equation}
  \label{eq:9}
  g_{ab}(x)=[e^{\sigma(x)}\hat{g}_{ab}]^{x^a\to f^a (x)}
\end{equation}
For surfaces of higher genus there are discrete families of equivalent metrics.

@@

!!!!!!!!!

Why it is so? How this definition is connected with another definitions and properties of conformal field theory?

!!!!!!!!!

@@

!!!! !!!! !!!! !!!! 
Be more elaborate here.
!!!!!!!

...
\section{Conformal Field Theory}
\label{sec:cft}

\subsection{General properties}
\label{sec:general-properties}

@@

\subsection{Wess-Zumino-Novikov-Witten models}
\label{sec:WZNW}

Let's start with non-linear $\sigma$-model.
\begin{equation}
  \label{eq:48}
  S_0=\frac{1}{4a^2}\int d^2x\; Tr' (\partial^{\mu}g^{-1}\partial_{\mu}g)
\end{equation}
Here $a^2>0$ is a positive parameter, $g(x)\in G$ - field on Lie group $G$, which we will assume to be semi-simple and by $Tr'$ we have denoted representation independent trace of Lie algebra $\mathfrak{g} $
\begin{equation}
  \label{eq:70}
  Tr'=\frac{1}{x_{rep}}Tr
\end{equation}
$x_{rep}$ is the Dynkin index of the representation.

In this model the conformal invariance is lost at the quantum level, as it can be seen along the lines of previous sections.
!!! Show it by example !!!

Also we can see that holomorphic and antiholomorphic currents aren't conserved individually. The equations of motion are
\begin{equation}
  \label{eq:71}
  \partial^{\mu}(g^{-1}\partial_{\mu}g)=0
\end{equation}
The currents are
\begin{equation}
  \label{eq:currents}
  J_{\mu}=g^{-1}\partial_{\mu}g
\end{equation}
or in the complex coordinates
\begin{equation}
  \label{eq:74}
  \begin{matrix}
    & \tilde{J}_z=g^{-1}\partial_z g, & \tilde{J}_{\bar{z}}=g^{-1}\partial_{\bar{z}}g\\
    & \partial_z \tilde{J}_{\bar{z}}+\partial_{\bar{z}}\tilde{J}_z=0 & \\
  \end{matrix}
\end{equation}
The terms of the equation of motion can not be null separately because $\partial_{\mu}(\epsilon^{\mu\nu}J_{\nu})\neq 0$.

So we add WZ-term to the action and redefine the currents in more appropriate way
\begin{equation}
  \label{eq:72}
  J_z=\partial_z g\;g^{-1}, \qquad J_{\bar{z}}=g^{-1}\partial{\bar z}g
\end{equation}

Wess-Zumino term has the following form
\begin{equation}
  \label{eq:73}
\Gamma=  - \frac{i }{24\pi} \int_{B}\epsilon_{ijk} Tr'\left(
    \tilde g^{-1}\frac{\partial \tilde g}{\partial y^i}
      \tilde g^{-1}\frac{\partial \tilde g}{\partial y^j}
      \tilde g^{-1}\frac{\partial \tilde g}{\partial y^k}\right) d^3y
\end{equation}

This term is defined on the three-dimensional manifold $B$ whose boundary is original two-dimensional space. By $\tilde{g}$ we have denoted the extension of the field $g$ to $B$. This extension isn't unique. In a compactified three-dimensional space a compact two-dimensional manifold delimits two three-dimensional manifolds. The difference between this two choices $\Delta\Gamma$ is given by the right hand side of (\ref{eq:73}) with the integration range extended over the whole compact three-dimensional space. Since the latter is topologically equivalent to the three-sphere we have
\begin{equation}
  \label{eq:75}
\Delta\Gamma=  - \frac{i }{24\pi} \int_{S^3}\epsilon_{ijk} Tr'\left(
    \tilde g^{-1}\frac{\partial \tilde g}{\partial y^i}
      \tilde g^{-1}\frac{\partial \tilde g}{\partial y^j}
      \tilde g^{-1}\frac{\partial \tilde g}{\partial y^k}\right) d^3y
\end{equation}
$\Delta\Gamma$ is defined modulo $2\pi i$, so the Euclidean functional integral with weight $exp(-\Gamma)$ is well-defined. Clearly, any coupling constant multiplying this term must be an integer.

We then consider the action
\begin{equation}
  \label{eq:76}
  S=S_0+k\Gamma
\end{equation}
where $k$ is an integer. 
The equation of motion for the full action (\ref{eq:76}) is
\begin{equation}
  \label{eq:77}
  \partial^{\mu}(g^{-1}\partial_{\mu}g)+\frac{a^2 ik}{4\pi}\epsilon_{\mu\nu}\partial^{\mu}(g^{-1}\partial^{\nu}g)=0
\end{equation}
It can be rewritten using complex coordinates as
\begin{equation}
  \label{eq:78}
  (1+\frac{a^2 k}{4\pi})\partial_z(g^{-1}\partial_{\bar z}g)+(1-\frac{a^2 k}{4\pi})\partial_{\bar z}(g^{-1}\partial_z g)=0
\end{equation}
We see that when $a^2=\frac{4\pi}{k}$ we have the desired conservation law
\begin{equation}
  \label{eq:79}
  \partial_z(g^{-1}\partial{\bar z}g)=0
\end{equation}
For currents
\begin{equation}
  \label{eq:4}
  \partial_{\bar z}J=0,\quad \partial_z \bar J=0
\end{equation}

The solution of classical equation is
\begin{equation}
  \label{eq:80}
  g(z,\bar z)=f(z)\bar f(\bar z)
\end{equation}
for arbitrary functions $f(z)$ and $\bar f (\bar z)$.

The separate conservation of the currents $J_z,\; J_{\bar z}$ implies the invariance of the action under the transformations
\begin{equation}
  \label{eq:81}
   g(z,\bar z)\to \Omega(z)g(z,\bar z)\bar \Omega^{-1}(\bar z)
\end{equation}
where $\Omega,\;\bar \Omega \in G$. So we have local $G(z)\times G(\bar z)$ invariance.

In order to move analysis to quantum level we rescale the currents
\begin{equation}
  \label{eq:82}
  J(z)\equiv -k \partial_zg g^{-1}\quad \bar J(\bar z)=k g^{-1}\partial_{\bar z}g
\end{equation}
So the variation of the action under infinitesimal transformation $\Omega=1+\omega,\; \bar \Omega =1+\bar \omega$ is
\begin{equation}
  \label{eq:83}
  \delta_{\omega,\bar\omega}S=\frac{i}{4\pi}\oint dz Tr' (\omega(z)J(z))-\frac{i}{4\pi}\oint d\bar z Tr'(\bar\omega(\bar z)\bar J(\bar z))
\end{equation}
Expanding the currents as
\begin{equation}
  \label{eq:85}
  \begin{aligned}
    J=\sum J^a t^a,\bar J=\sum \bar J^a t^a \\
    \omega=\sum \omega^a t^a\\
  \end{aligned}
\end{equation}
we get
\begin{equation}
  \label{eq:86}
  \delta_{\omega,\bar \omega}S=-\frac{1}{2\pi i}\oint dz \sum\omega^a J^a+\frac{1}{2\pi i} \oint d\bar z \sum \bar \omega^a \bar J^a
\end{equation}
We can also obtain Ward identities $\delta\left< X\right>=\left<(\delta S)X\right>$
\begin{equation}
  \label{eq:87}
  \delta_{\omega,\bar \omega}\left< X \right>=-\frac{1}{2\pi i}\oint dz \sum\omega^a \left< J^a X\right>+
  \frac{1}{2\pi i} \oint d\bar z \sum \bar \omega^a \left< \bar J^a X\right>
\end{equation}
Then for the currents we have
\begin{equation}
  \label{eq:88}
  \delta_{\omega}J=[\omega,J]-k\partial_z\omega,\quad \delta_{\omega}J^a=\sum i f_{abc}\omega^b J^c-k\partial_z\omega^a
\end{equation}
The operator product expansion for currents has the form
\begin{equation}
  \label{eq:89}
  J^a(z) J^b(w) \sim \frac{k\delta_{ab}}{(z-w)^2}+\sum i f_{abc}\frac{J^c(w)}{(z-w)}
\end{equation}
Expanding currents into series we get
\begin{equation}
  \label{eq:90}
  \begin{aligned}
    J^a(z)=\sum_{n\in \mathbb Z}z^{n-1}J^a_n\\    
    \left[J^a_n,J^b_m\right]=\sum_c i f^{abc}J^c_{n+m}+kn\delta^{ab}\delta_{n+m,0}  
  \end{aligned}
\end{equation}
We see that components of currents realize affine Lie algebra $\hat g$.

Energy-momentum tensor is introduced with Sugawara construction as the sum of normal-ordered products of current components
\begin{equation}
  \label{eq:6}
  T(z)=\frac{1}{2(k+h^v)}\sum_a N(J^a J^a)(z)
\end{equation}
Here $h^v$ is the dual Coxeter number.

Energy-momentum tensor can be expanded into modes $L_n$
\begin{equation}
  \label{eq:91}
  L_n=\frac{1}{2(k+h^v)}\sum_a\sum_m:J^a_m J^a_{n-m}:
\end{equation}
The commutation relations for the modes $L_n$ are
\begin{equation}
  \label{eq:92}
  \begin{aligned}
    \left[L_n,L_m\right]=(n-m)L_{n+m}+\frac{c}{12}(n^3-n)\delta_{n+m,0}\\
    \left[L_n,J^a_m\right]=-mJ^a_{n+m}
  \end{aligned}
\end{equation}

So the Sugawara construction is the way to embed Virasoro algebra into the universal enveloping of affine Lie algebra.

Complete chiral algebra of WZW-model is equal to semidirect product $Vir\ltimes \hat g$

The primary fields are the fields that transforms covariantly with respect to $G(z)\times G(\bar z)$, such as $g(z,\bar z)$. In terms of operator product expansion this property is reformulated as
\begin{equation}
  \label{eq:84}
  \begin{aligned}
    J^a(z)g(w,\bar w)\sim \frac{-t^a g(w,\bar w)}{(z-w)}\\
    \bar J^a(z)g(w,\bar w)\sim \frac{ g(w,\bar w)t^a}{(z-w)}
  \end{aligned}
\end{equation}
Any field $\phi_{\lambda,\mu}$ transforming covariantly with respect to some representation specified by $\lambda$ in holomorphic and $\mu$ in antiholomorphic sector, will be a WZW primary field.

In terms of modes this property can be written as
\begin{equation}
  \label{eq:93}
  \begin{aligned}
    & (J_0^a \phi_{\lambda})=-t^a_{\lambda}\phi_{\lambda}\\
    & (J^a_n\phi_{\lambda})=0\quad \mbox{for}\; n>0\\
  \end{aligned}
\end{equation}
We can associate the state $\left|\phi_{\lambda}\right>$ to the field $\phi_{\lambda}$
  \begin{equation}
    \label{eq:94}
    \phi_{\lambda}(0)=\left|\phi_{\lambda}\right>
  \end{equation}
Then the conditions (\ref{eq:93}) for WZW primary field are
\begin{equation}
  \label{eq:95}
  \begin{aligned}
    & J_0^a\left|\phi_{\lambda}\right>=-t^a_{\lambda}\left|\phi_{\lambda}\right>\\
    & J^a_n\left|\phi_{\lambda}\right>=0 \quad \mbox{for}\; n>0 \\
  \end{aligned}
\end{equation}
It is easy to see that the action of the Virasoro generators is
\begin{equation}
  \label{eq:96}
  L_0\left|\phi_{\lambda}\right>=\frac{1}{2(k+h^v)}\sum_aJ^a_0J^a_0\left|\phi_{\lambda}\right>=\frac{(\lambda,\lambda+2\rho)}{2(k+h^v)}\left|\phi_{\lambda}\right>
\end{equation}
Here we used the explicit form of the quadratic Casimir operator eigenvalue.

Primary fields live in integrable finite-dimensional representations since infinite-dimensional and non-integrable fields are decoupled from the correlation functions.

>>>>

\documentclass[a4paper,12pt]{article}
\usepackage[unicode,verbose]{hyperref}
\usepackage{amsmath,amssymb,amsthm} \usepackage{pb-diagram}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage[russian]{babel}
\usepackage{cmap}
\usepackage{graphicx}
\pagestyle{plain}
\theoremstyle{definition} \newtheorem{Def}{Definition}
\title{WZW-модели и теория Черна-Саймонса}
\author{Антон Назаров\\ antonnaz@gmail.com}
\begin{document}
\maketitle
\begin{abstract}
WZW-модель - это сигма-модель с дополнительным членом Весса-Зумино, связанным с топологией. Существует связь таких моделей с пределом бесконечного числа цветов в калибровочных теориях. WZW-модели также связаны с моделями топологической теории поля - моделями Черна-Саймонса-Виттена в трех измерениях. Квантовые состояния в теории Черна-Саймонса с уровнем $k$ ($SU(2)_k$) эквивалентны квантовым состояниям WZW-модели, связанной с представлениями уровня $k$ аффинной алгебры Ли $\widehat{su(2)}$.
\end{abstract}

\section{Конформная инвариантность}
\label{sec:conformal-invariance}

Откуда берется конформная инвариантность в физике конденсированного состояния и в теории струн?

Конформные преобразования:

    \begin{align} & M_{\mu\nu} \equiv i(x_\mu\partial_\nu-x_\nu\partial_\mu) \,, \\ &P_\mu \equiv-i\partial_\mu \,, \\ &D \equiv-ix_\mu\partial^\mu \,, \\ &K_\mu \equiv i(x^2\partial_\mu-2x_\mu x_\nu\partial^\nu) \,, \end{align}

Конформная инвариантность тесно связана с масштабной инвариантностью и иногда даже прямо из нее следует. В физике конденсированного состояния масштабная инвариантность имеет место при фазовом переходе (стремление параметра порядка к бесконечности). Флуктуации при фазовом переходе конформно-инвариантны, что позволяет классифицировать фазовые переходы в терминах конформной теории поля \cite{Polyakov:1970xd}, \cite{belavin1984ics}.

В теории струн действие инвариантно относительно репараметризаций мировой поверхности, в том числе конформных преобразований. 
Однако эта инвариантность не может быть сохранена при квантовании, поэтому возникает конформная аномалия и центральный заряд алгебры Вирасоро.

В двумерной теории поля локальная конформная группа бесконечномерна, хотя глобальная конформная группа конечномерна. Например, в конформной теории поля на сфере Римана глобальные конформные преобразования образуют группу Мебиуса $PSL(2,\mathbb{C})$. Алгебра же инфинитезимальных преобразований в классической теории --- это алгебра Витта, при квантовании возникает конформная аномалия, что соответствует центральному расширению алгебры. Так появляется алгебра Вирасоро.

\subsection{Конформная инвариантность в теории струн}
\label{sec:conformal-invariance-strings}

Действие свободной релятивистской частицы --- это длина траектории.
\begin{equation}
  \label{eq:1}
  \begin{split}
  x^{\mu}(0)=x^{\mu}_0,\; x^{\mu}(1)=x^{\mu}_1\\
  S=\int_{0}^{1} \sqrt{\left(\frac{dx^{\mu}}{d\tau}^2\right)}d\tau
\end{split}
\end{equation}
Оно Пуанкаре-инвариантно $x^{\mu}\to A^{\mu}_{\nu}x^{\nu}+ B^{\mu}$ и не зависит от репараметризаций $\tau\to f(\tau),\; x^{\mu}(\tau)\to x^{\mu}(f(\tau))$.

Аналогично, струнное действие дается площадью мировой поверхности
\begin{equation}
  \label{eq:2}
  \begin{split}
    X^{\mu}(x^1,x^2),\; \mu=1\dots d\\
    S_N [X^{\mu}(x^1,x^2)]=\int \sqrt{det \frac{\partial X^{\mu}}{\partial x^a}\frac{\partial X^{\mu}}{\partial x^b}} dx^1 dx^2
  \end{split}
\end{equation}
Такое действие называется действием Намбу. Оно инвариантно относительно преобразований Лоренца и репараметризаций.

Амплитуда перехода для струны дается следующей формулой:
\begin{equation}
  \label{eq:3}
  A=\int {\cal D}X^{\mu}(x^1,x^2)e^{-S_N[X^{\mu}(x^1,x^2)]}
\end{equation}
Сложность, как всегда, состоит в определении меры интегрирования. Она плохо определена как и в большинстве случаев, так что теория струн здесь не выделяется.

Мы хотим сделать замену ``переменных интегрирования'', чтобы избавиться от произвола в выборе параметризации мировой поверхности.

Теория должна быть инвариантна относительно репараметризаций:
\begin{equation}
  \label{eq:4}
  \begin{split}
    x^a \to f^a (x^1,x^2)\\
    X^{\mu}(x^1,x^2)\to \tilde{X}^{\mu}(x^1,x^2)=X^{\mu}(f^1(x^1,x^2),f^2(x^1,x^2))
  \end{split}
\end{equation}
Но в действии Намбу эта инвариантность не видна, поэтому мы перейдем к эквивалентному действию Полякова:
\begin{equation}
  \label{eq:5}
  \begin{split}
    S_P [g^{ab}(x^1,x^2), X^{\mu}(x^1,x^2)]=\int g^{ab}(x)\partial_a X^{\mu} \partial_b X^{\mu}\sqrt{g} \;d^2 x\\
    g=\det g_{ab}
  \end{split}
\end{equation}

Действия Полякова и Намбу эквивалентны в том смысле, что уравнения движения совпадают после исключения вспомогательного поля $g_{ab}$.
\begin{equation}
  \label{eq:6}
  \delta_{X}S_N=0 \Longleftrightarrow \left\{
      \begin{aligned}
        \delta_X S_P=0\\
        \delta_g S_P=0
      \end{aligned}
      \right.
\end{equation}
Действие Полякова инвариантно относительно преобразований Лоренца и репараметризаций мировой поверхности:
\begin{equation}
  \label{eq:7}
  \begin{split}
    x^a \to f^a (x^1,x^2)\\
    X^{\mu}(x^1,x^2)\to \tilde{X}^{\mu}(x^1,x^2)=X^{\mu}(f^1(x^1,x^2),f^2(x^1,x^2))\\
    g_{ab}(x)\to \tilde{g}_{ab}(x)=\frac{\partial f^{a_1}}{\partial x^{a}}\frac{\partial f^{b_1}}{\partial x^b} g_{a_1 b_1}(x)
  \end{split}  
\end{equation}
Кроме того, есть дополнительная симметрия, которая называется вейлевской или конформной инвариантностью:
\begin{equation}
  \label{eq:8}
  \begin{split}
    g_{ab}(x)\to e^{\sigma(x)}g_{ab}(x)\\
    g\to e^{\sigma} g\\
    g^{ab}\to e^{-\sigma} g^{ab}\\
    S_P[e^{\sigma(x)} g_{ab}(x),X^{\mu}] = S_P[g_{ab}(x),X^{\mu}] 
  \end{split}
\end{equation}

То есть у нас довольно большая свобода --- мы можем, например, получить произвольную метрику на сфере из постоянной метрики путем репараметризации и преобразования Вейля
\begin{equation}
  \label{eq:9}
  g_{ab}(x)=[e^{\sigma(x)}\hat{g}_{ab}]^{x^a\to f^a (x)}
\end{equation}
Для поверхностей большего рода существуют дискретные семейства эквивалентных метрик.

@@

@@

@@

\subsection{Глобальная конформная инвариантность и уравнения Книжника-Замолодчикова}
\label{sec:knizhnik-zamolodchikov}

\subsection{Модулярные преобразования}
\label{sec:modular}

Конформная инвариантность в квантовой и классической теории поля.
Глобальная конформная инвариантность.
Уравнения Книжника-Замолодчикова.
Локальная конформная инвариантность. Бесследовость тензора энергии-импульса.
Голоморфная факторизация.

\section{Модели Весса-Зумино-Новикова-Виттена}
\label{sec:WZNW}

Связь WZW-моделей с топологией, обоснование с точки зрения предела калибровочной теории.

Стартуем с нелинейной $\sigma$-модели.
\begin{equation}
  \label{eq:48}
  S_0=\frac{1}{4a^2}\int d^2x\; Tr' (\partial^{\mu}g^{-1}\partial_{\mu}g)
\end{equation}
Здесь $a^2>0$ - положительный параметр, $g(x)\in G$ - поле со значениями в группе $G$, которую мы будем считать полупростой, а через $Tr'$ мы обозначили след, не зависящий от выбора представления алгебры Ли $\mathfrak{g} $
\begin{equation}
  \label{eq:70}
  Tr'=\frac{1}{x_{rep}}Tr
\end{equation}
$x_{rep}$ - индекс Дынкина представления.

В нелинейной $\sigma$-модели конформная инвариантность теряется на квантовом уровне.
Голоморфный и антиголоморфный токи не сохраняются по отдельности. Уравнения движения имеют вид    
\begin{equation}
  \label{eq:71}
  \partial^{\mu}(g^{-1}\partial_{\mu}g)=0
\end{equation}
Токи задаются выражением
\begin{equation}
  \label{eq:currents}
  J_{\mu}=g^{-1}\partial_{\mu}g
\end{equation}
или в комплексных координатах
\begin{equation}
  \label{eq:74}
  \begin{matrix}
    & \tilde{J}_z=g^{-1}\partial_z g, & \tilde{J}_{\bar{z}}=g^{-1}\partial_{\bar{z}}g\\
    & \partial_z \tilde{J}_{\bar{z}}+\partial_{\bar{z}}\tilde{J}_z=0 & \\
  \end{matrix}
\end{equation}
Члены в уравнении движения не разделяются, так как  $\partial_{\mu}(\epsilon^{\mu\nu}J_{\nu})\neq 0$.

Поэтому мы добавляем член Весса-Зумино к действию и переопределяем токи:
\begin{equation}
  \label{eq:72}
  J_z=\partial_z g\;g^{-1}, \qquad J_{\bar{z}}=g^{-1}\partial{\bar z}g
\end{equation}

Член Весса-Зумино имеет вид
\begin{equation}
  \label{eq:73}
\Gamma=  - \frac{i }{24\pi} \int_{B}\epsilon_{ijk} Tr'\left(
    \tilde g^{-1}\frac{\partial \tilde g}{\partial y^i}
      \tilde g^{-1}\frac{\partial \tilde g}{\partial y^j}
      \tilde g^{-1}\frac{\partial \tilde g}{\partial y^k}\right) d^3y
\end{equation}

Он определен на трехмерном многообразии $B$, ограниченном исходным двумерным пространством. 
Через $\tilde{g}$ мы обозначили продолжение поля  $g$ на $B$. Такое продолжение не единственно. В компактифицированном трехмерном пространстве компактное двумерное многообразие разделяет два трехмерных многообразия. Разность значений члена Весса-Зумино  $\Delta\Gamma$ на этих многообразиях дается правой частью уравнения (\ref{eq:73}) с интегралом, продолженным на все компактное трехмерное пространство. Так как оно топологически эквивалентно три-сфере, получаем
\begin{equation}
  \label{eq:75}
\Delta\Gamma=  - \frac{i }{24\pi} \int_{S^3}\epsilon_{ijk} Tr'\left(
    \tilde g^{-1}\frac{\partial \tilde g}{\partial y^i}
      \tilde g^{-1}\frac{\partial \tilde g}{\partial y^j}
      \tilde g^{-1}\frac{\partial \tilde g}{\partial y^k}\right) d^3y
\end{equation}
$\Delta\Gamma$ определен по модулю $2\pi i$, поэтому Евклидов функциональный интеграл с весом  $exp(-\Gamma)$ хорошо определен. Значит константа связи, умножаемая на этот член, должна быть целочисленной.

Теперь мы рассматриваем действи
\begin{equation}
  \label{eq:76}
  S=S_0+k\Gamma
\end{equation}
где $k$ - целое.
Уравнение движения для полного действия (\ref{eq:76}):
\begin{equation}
  \label{eq:77}
  \partial^{\mu}(g^{-1}\partial_{\mu}g)+\frac{a^2 ik}{4\pi}\epsilon_{\mu\nu}\partial^{\mu}(g^{-1}\partial^{\nu}g)=0
\end{equation}
В комплексных координатах оно записывается в виде
\begin{equation}
  \label{eq:78}
  (1+\frac{a^2 k}{4\pi})\partial_z(g^{-1}\partial_{\bar z}g)+(1-\frac{a^2 k}{4\pi})\partial_{\bar z}(g^{-1}\partial_z g)=0
\end{equation}
Видно, что при $a^2=\frac{4\pi}{k}$ у нас имеются интересующие нас законы сохранения
\begin{equation}
  \label{eq:79}
  \partial_z(g^{-1}\partial{\bar z}g)=0
\end{equation}
Для токов
\begin{equation}
  \label{eq:4}
  \partial_{\bar z}J=0,\quad \partial_z \bar J=0
\end{equation}

Решение классического уравнения движения имеет вид
\begin{equation}
  \label{eq:80}
  g(z,\bar z)=f(z)\bar f(\bar z)
\end{equation}
при произвольных $f(z)$ и $\bar f (\bar z)$.

Сохранение по отдельности токов $J_z,\; J_{\bar z}$ приводит к инвариантности действия при преобразованиях
\begin{equation}
  \label{eq:81}
   g(z,\bar z)\to \Omega(z)g(z,\bar z)\bar \Omega^{-1}(\bar z)
\end{equation}
где $\Omega,\;\bar \Omega \in G$. То есть мы получили локальную $G(z)\times G(\bar z)$-инвариантность.

Для перехода к квантовому случаю мы переопределяем токи
\begin{equation}
  \label{eq:82}
  J(z)\equiv -k \partial_zg g^{-1}\quad \bar J(\bar z)=k g^{-1}\partial_{\bar z}g
\end{equation}
Тогда вариация действия при инфинитезимальных преобразованиях $\Omega=1+\omega,\; \bar \Omega =1+\bar \omega$ дается выражением
\begin{equation}
  \label{eq:83}
  \delta_{\omega,\bar\omega}S=\frac{i}{4\pi}\oint dz Tr' (\omega(z)J(z))-\frac{i}{4\pi}\oint d\bar z Tr'(\bar\omega(\bar z)\bar J(\bar z))
\end{equation}
Раскладывая токи
\begin{equation}
  \label{eq:85}
  \begin{aligned}
    J=\sum J^a t^a,\bar J=\sum \bar J^a t^a \\
    \omega=\sum \omega^a t^a\\
  \end{aligned}
\end{equation}
получаем
\begin{equation}
  \label{eq:86}
  \delta_{\omega,\bar \omega}S=-\frac{1}{2\pi i}\oint dz \sum\omega^a J^a+\frac{1}{2\pi i} \oint d\bar z \sum \bar \omega^a \bar J^a
\end{equation}
Мы также получили тождества Уорда $\delta\left< X\right>=\left<(\delta S)X\right>$
\begin{equation}
  \label{eq:87}
  \delta_{\omega,\bar \omega}\left< X \right>=-\frac{1}{2\pi i}\oint dz \sum\omega^a \left< J^a X\right>+
  \frac{1}{2\pi i} \oint d\bar z \sum \bar \omega^a \left< \bar J^a X\right>
\end{equation}
Для токов имеем
\begin{equation}
  \label{eq:88}
  \delta_{\omega}J=[\omega,J]-k\partial_z\omega,\quad \delta_{\omega}J^a=\sum i f_{abc}\omega^b J^c-k\partial_z\omega^a
\end{equation}
Операторное разложение для токов имеет вид 
\begin{equation}
  \label{eq:89}
  J^a(z) J^b(w) \sim \frac{k\delta_{ab}}{(z-w)^2}+\sum i f_{abc}\frac{J^c(w)}{(z-w)}
\end{equation}
Раскладывая токи в ряд, получаем
\begin{equation}
  \label{eq:90}
  \begin{aligned}
    J^a(z)=\sum_{n\in \mathbb Z}z^{n-1}J^a_n\\    
    \left[J^a_n,J^b_m\right]=\sum_c i f^{abc}J^c_{n+m}+kn\delta^{ab}\delta_{n+m,0}  
  \end{aligned}
\end{equation}
Теперь мы видим, что компоненты токов образуют аффинную алгебру Ли $\hat g$.

Тензор энергии-импульса вводится при помощи конструкции Сугавары как сумма нормально упорядоченных компонент токов
\begin{equation}
  \label{eq:6}
  T(z)=\frac{1}{2(k+h^v)}\sum_a N(J^a J^a)(z)
\end{equation}
Здесь $h^v$ - дуальное число Кокстера.

Тензор энергии-импульса можно разложить на моды $L_n$
\begin{equation}
  \label{eq:91}
  L_n=\frac{1}{2(k+h^v)}\sum_a\sum_m:J^a_m J^a_{n-m}:
\end{equation}
Тогда коммутационные соотношения для мод  $L_n$ имеют вид
\begin{equation}
  \label{eq:92}
  \begin{aligned}
    \left[L_n,L_m\right]=(n-m)L_{n+m}+\frac{c}{12}(n^3-n)\delta_{n+m,0}\\
    \left[L_n,J^a_m\right]=-mJ^a_{n+m}
  \end{aligned}
\end{equation}

Таким образом, конструкция Сугавары --- это способ вложения алгебры Вирасоро в универсальную обертывающую аффинной алгебры Ли $\hat{g}$

Полная киральная алгебра модели Весса-Зумино-Виттена равна полупрямому произведению $Vir\ltimes \hat g$

Примарными называются поля, которые преобразуются ковариантно под действием $G(z)\times G(\bar z)$, как $g(z,\bar z)$. В терминах операторного разложения это свойство переформулируется следующим образом:
\begin{equation}
  \label{eq:84}
  \begin{aligned}
    J^a(z)g(w,\bar w)\sim \frac{-t^a g(w,\bar w)}{(z-w)}\\
    \bar J^a(z)g(w,\bar w)\sim \frac{ g(w,\bar w)t^a}{(z-w)}
  \end{aligned}
\end{equation}
Любое поле $\phi_{\lambda,\mu}$, преобразующееся ковариантно по отношению к некоторому представлению, заданному весом  $\lambda$ в голоморфном секторе и весом $\mu$ в антиголоморфном, является примарным полем WZW-модели.

В модах это свойство записывается в виде
\begin{equation}
  \label{eq:93}
  \begin{aligned}
    & (J_0^a \phi_{\lambda})=-t^a_{\lambda}\phi_{\lambda}\\
    & (J^a_n\phi_{\lambda})=0\quad \mbox{для}\; n>0\\
  \end{aligned}
\end{equation}
Мы можем сопоставить состояние $\left|\phi_{\lambda}\right>$ полю $\phi_{\lambda}$
  \begin{equation}
    \label{eq:94}
    \phi_{\lambda}(0)=\left|\phi_{\lambda}\right>
  \end{equation}
Тогда условия  (\ref{eq:93}) для примарных полей  дают
\begin{equation}
  \label{eq:95}
  \begin{aligned}
    & J_0^a\left|\phi_{\lambda}\right>=-t^a_{\lambda}\left|\phi_{\lambda}\right>\\
    & J^a_n\left|\phi_{\lambda}\right>=0 \quad \mbox{для}\; n>0 \\
  \end{aligned}
\end{equation}
Легко видеть, что действие генераторов алгебры Вирасоро имеет вид
\begin{equation}
  \label{eq:96}
  L_0\left|\phi_{\lambda}\right>=\frac{1}{2(k+h^v)}\sum_aJ^a_0J^a_0\left|\phi_{\lambda}\right>=\frac{(\lambda,\lambda+2\rho)}{2(k+h^v)}\left|\phi_{\lambda}\right>
\end{equation}
Здесь использовано явное выражение для собственных значений квадратичного оператора Казимира.

Примарные поля живут в интегрируемых конечномерных представлениях, так как бесконечномерные и неинтегрируемые поля отщепляются в корреляционных функциях.
